Staged Abstract Interpreters

### 2018-11-02 paper/10_intro.tex
fix paper

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter yields a compiler. \note{TR: not true!!}
Multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} is a popular way to write 
generic programs that and generates specialized and efficient programs;  \note{TR: connection from Futamura to MSP doesn't flow well}
the generated program then can be executed at next stage. 
As an approach to partial evaluation, MSP requires explicit staging annotations from the implementers.
The staging annotations controls which part of the program should be specialized, and are considered 
as manual binding-time analysis. On the other hand, partial evaluation \cite{DBLP:books/daglib/0072559} 
can be considered as a two-stage (static or dynamic) instance of MSP but analyzing binding-time automatically. 
\note{TR: are last two sentences relevant? perhaps rephrase in the direction of ``here's who to implement
the futamura projection idea in practice''}

@@

\begin{itemize}
\item Starting from a definitional abstract interpreter for a higher-order functional language,
  whose syntax and semantics are described in section \ref{bg_lang},
  we show how to turn it into a staged abstract interpreter by adding stage annotations with
  the Lightweight Modular Staging (LMS) framework. %, a library-based approach to code generation. 
  We later extend the staged abstract interpreter to support imperative language features 
  (section \ref{cases_imp}), such as states, loops and exceptions.
\item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
  but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
  programs. \todo{more on this point.}
\item To demonstrate the feasibility and applicability of the proposed approach, we conduct three case studies in section \ref{cases_study}. 
  First we revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587}, which is a closure generation 
  technique applied on 0-CFA. We show that it can be understood and easily implemented as an instance of 
  our staged abstract interpreter. 
  Second, we demonstrate sophisticated control-flow analysis techniques, such as adaptive context/flow/path-sensitive 
  analysis and store-widening can be integrated with staged abstract interpreters without losing any precision.
  Finally, we show that in an imperative language, the idea can be applied to an abstract interpreter with
  an abstract domain that performs numerical analysis.
  % (Optional: taint/information flow analysis?)
\item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
  for control-flow analysis. We compare against with the unstaged version, a small-step version (unstaged), 
  and an optimizing small-step version.
\end{itemize}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to the input program yields 
an equivalent executable.
Multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} is a popular way to write 
generic programs that and generates specialized and efficient programs;  \note{TR: connection from Futamura to MSP doesn't flow well}
the generated program then can be executed at next stage. 
As an approach to partial evaluation, MSP requires explicit staging annotations from the implementers.
The staging annotations controls which part of the program should be specialized, and are considered 
as manual binding-time analysis. On the other hand, partial evaluation \cite{DBLP:books/daglib/0072559} 
can be considered as a two-stage (static or dynamic) instance of MSP but analyzing binding-time automatically. 
\note{TR: are last two sentences relevant? perhaps rephrase in the direction of ``here's who to implement
the futamura projection idea in practice''}

@@

\begin{itemize}
\item Starting from a definitional abstract interpreter for a higher-order functional language,
  whose syntax and semantics are described in section \ref{bg_lang},
  we show how to turn it into a staged abstract interpreter (SAI) by adding stage annotations. 
  The design and implementation of SAI using Scala and Lightweight Modular Staging (LMS) framework is 
  described in section \ref{sai}.
  We later extend the staged abstract interpreter to support imperative language features 
  (section \ref{cases_imp}), such as states, loops and exceptions.
\item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
  but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
  programs. \todo{more on this point.}
\item To demonstrate the feasibility and applicability of the proposed approach, we conduct three case studies in section \ref{cases_study}. 
  First we revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587}, which is a closure generation 
  technique applied on 0-CFA. We show that it can be understood and easily implemented as an instance of 
  our staged abstract interpreter. 
  Second, we demonstrate sophisticated control-flow analysis techniques, such as adaptive context/flow/path-sensitive 
  analysis and store-widening can be integrated with staged abstract interpreters without losing any precision.
  Finally, we show that in an imperative language, the idea can be applied to an abstract interpreter with
  an abstract domain that performs numerical analysis.
  % (Optional: taint/information flow analysis?)
\item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
  for control-flow analysis. We compare against with the unstaged version, a small-step version (unstaged), 
  and an optimizing small-step version.
\end{itemize}

### 2018-11-02 paper/20_bg.tex
fix paper

@@

\subsection{The Language} \label{bg_lang}

syntax, higher-order functional, variables, lambda, app, fix, let, arith, if0

concrete interpreter.

@@

big-step abstract interpreter using delimited control operators \cite{Wei:2018:RAA:3243631.3236800}.

monadic big-step abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17}.

\subsection{Multi-Stage Programming}

general idea; LMS framework in scala. MetaOCaml?

@@

>>>>

\newcommand{\TLang}{$L_\lambda$}

@@

In this section, we first described the language we will model and analyze, and present a unstaged concrete 
interpreter for it. Then, starting from this unstaged concrete interpreter, we proceed in two directions:
by abstracting it we obtain an unstaged abstract interpreter; by staging it we obtain a staged concrete
interpreter.

\subsection{The Target Language} \label{bg_lang}

As higher-order functions had been widely introduced in mainstream programming languages \todo{cite}, in this paper, 
we target a tiny higher-order functional language \TLang, which is based on direct-style lambda-calculus, 
and with the support of numbers, arithmetics, recursions and conditionals.
In section \ref{cases_imp}, we will see adding more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, the static semantics and
type system will be omitted. We also assume all the input programs are well-typed.
The abstract syntax is described below:

\begin{lstlisting}
abstract class Expr
case class Lit(i: Int)
case class Var(x: String)
case class Lam(x: String, e: Expr)
case class App(e1: Expr, e2: Expr)
case class AOp(op: String, e1: Expr, e2: Expr)
case class Rec(x: String, e: Expr, body: Expr)
case class If0(cnd: Expr, thn: Expr, els: Expr)
\end{lstlisting}

We give the concrete semantics of \TLang by showing a big-step definitional interpreter for it.
But before doing that, let's first see a general interface for interpreters, whether concrete or abstract, 
staged or unstaged. The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Ident},
\texttt{Addr} and etc., which will be instantiated concretely later. The abstract types \texttt{Env} and \texttt{Store}
are required to conform their specifications, \texttt{EnvSpec} and \texttt{StoreSpec}, which define 
the operations such as lookup and update on them, nevertheless the detail of them are omitted.  
The store component is necessary when modeling side effects such as mutations. 
The trait also requires the implementators provide the initial values of environment and store.

%TODO: talk about alloc here?

\begin{lstlisting}
trait Semantics[R[_]] {
  type Ident;    type Addr;   type Value;    type Ans
  type Env <: EnvSpec[R, Ident, Addr, Env]
  type Store <: StoreSpec[R, Addr, Value, Store]
  val $\rho$0: Env;     val $\sigma$0: Store
  def alloc(x: Ident, $\sigma$: Store): R[Addr]
  type EvalFun = (Expr, Env, Store) => Ans
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans
  def fix(ev: EvalFun => EvalFun): EvalFun = 
    (e, $\rho$, $\sigma$) =>  ev(fix(ev))(e, $\rho$, $\sigma$)
  def eval_top(e: Expr): Ans = fix(eval)(e, $\rho$0, $\sigma$0)
}
\end{lstlisting}

The evaluator \texttt{eval} is a open-recursive function that takes one and three arguments: the first
one \texttt{ev} represents the recursive function itself; the rest three are the AST representing program 
will be executed, the environment that maps from identifiers to addresses, and the store that maps addresses to values.
The interpreters that take control, environment and store components also can be obtained by 
refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03, Wei:2018:RAA:3243631.3236800} a small-step CESK 
machine \cite{DBLP:conf/popl/FelleisenF87}.
We provide a function \texttt{fix} to close the recursion by providing the result of \texttt{fix(ev)} 
to argument \texttt{ev}. Finally, a default implementation for the top-level evaluation function \texttt{eval\_top} is defined.
As we will see later, the open-recursion style is beneficial when implementing abstract interpreter and 
staged abstract interpreter \todo{cite adi}.

Readers may notice an additional notation \texttt{[R[\_]]} afterward the trait definition,
which indicates \texttt{Semantics} takes a type parameter \texttt{R}, where \texttt{R} represents
a higher-kinded type and will used to specify the binding-time, i.e., staged or not. 
For the moment, \texttt{R} will be instantiated unstaged --- an identity type constructor: 
\texttt{type NoRep[A] = A}. \todo{where talk about stage poly?}

Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, the address space is infinite thus type \texttt{Int} is used here.
The final returned type \texttt{Ans} is a pair of \texttt{Value} and \texttt{Store}, where
the values are either closure or numbers.

\begin{lstlisting}
trait Concrete[R[_]] extends Semantics[R] {
  type Ident = String;         type Addr = Int
  type Ans = (Value, Store);   abstract class Value
  case class CloV($\lambda$: Lam, $\rho$: Env) extends Value
  case class NumV(i: Int) extends Value
}
\end{lstlisting}

Then we define evaluation function that dispatches differently for different expression.
Note that an allocation function is used when we need to update the store: 
in the concrete case, function \texttt{alloc} always returns a fresh address of the store.

\begin{lstlisting}
object NoRepConcreteEval extends Concrete[NoRep] {
  def alloc(x: Ident, $\sigma$: Store): Addr = $\sigma$.size + 1
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
    e match {
      case Lit(i) => (NumV(i), $\sigma$)
      case Var(x) => ($\sigma$($\rho$(x)), $\sigma$)
      case Lam(x, e) => (CloV(Lam(x,e), $\rho$), $\sigma$)
      case App(e1, e2) =>
        val (CloV(Lam(x,e), $\lambda\rho$), e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        val $\alpha$ = alloc(x, e2$\sigma$)
        ev(e, $\lambda\rho$ + (x -> $\alpha$), e2$\sigma$ + ($\alpha$ -> e2v))
      case If0(cnd, thn, els) => ...
      case AOp(op, e1, e2) => ...
      case Rec(x, e, body) => ...
    }
}
\end{lstlisting}

@@

\begin{lstlisting}
trait Abstract[R[_]] extends Semantics[R] {
  type Ident = String;         case class Addr(x: Ident)
  type Ans = (Value, Store);   type Value = Set[AbsValue]
  abstract class AbsValue
  case class CloV($\lambda$: Lam, $\rho$: Env) extends AbsValue
  case class NumV() extends AbsValue
}
\end{lstlisting}

Address just record the variable name, more sensitivity in section ...

other big-step abstract interpreter, 
big-step abstract interpreter using delimited control operators \cite{Wei:2018:RAA:3243631.3236800},
monadic big-step abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17},
arrow-based abstract interpreter \cite{Keidel:2018:CSP:3243631.3236767}.

\subsection{From Interpreter to Staged Interpreter}

\subsubsection{Multi-Stage Programming}
\todo{example, power?}
general idea; LMS framework in scala. MetaOCaml?

@@

\subsubsection{Staged Interpreters}
